On the spatial aggregation of ecosystem condition indicators

### 2025-02-04 manuscript/manuscript.qmd
update agg bias fig

For sparse data we are considering data where variable values are assigned to singular points or small areas.
In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information.
To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some generalisations.
One can assign values to areas via an aggregation functions used on data values that fall within a given ecosystem asset.
Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54] .
When doing this aggregation, we need to account for the spatial structure (spatial correlations) of the data (@sec-auto).
One way is to rely on design-based approaches (relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets.
In our experience, the choice of which area units to use when doing this step-wise aggregation is often not reflected much upon.
In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
However, the arbitrariness of this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; @nte-rel).
It is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
The MAUP can, however, lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].

@@

```{r fig-agg1}
#| echo: false
#| fig-cap: |
#|   Example showing the difference in indicator values (y-axis) obtained from taking the mean of two variable values (V1 and V1) that are on the original scale (i hat) or on the normalised indicator scale (i bar). Due to truncation at x100, the latter results in a comparatively lower indicator value. The solid coloured line represents the normalisation function f(V). V = variable value, i = indicator value. Modified from Rastsetter (1991).
#| out-width: 400

knitr::include_graphics("../img/aggregation_bias_truncations.PNG")

@@

```{r fig-agg2}
#| echo: false
#| fig-cap: |
#|   Same figure as above, but showing an example where the normalisation function includes a transformation step (see Figure 1 for definition) that anchors X60 to 0.6 on the indicator scale.
#| out-width: 400

knitr::include_graphics("../img/aggregation_bias_x60.PNG")
```

@@

In ecosystem condition accounting the aggregation of both variables and indicators is common practice and raises the issue of **aggregation bias** (@fig-agg1; @fig-agg2) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
This means that the spatial aggregation of metrics leads to a divergence in the type of information held in variables and in their associated indicators, to a point where they become qualitatively very different.
As an example, consider an indicator based on spatial information on the abundance of a large herbivore.
The ecosystem characteristic that this metric is meant to describe is "*Appropriate herbivore pressure, balancing overgrazing and shrubification"*.
The variable shows relatively high densities in half the area and low densities in the other half.
When spatially aggregating the variable, either by taking the mean density or the sum of individuals, will produce an estimate that is somewhere in the middle of the two regions.
The user will have no information about the carrying capacity for each of the regions, and will only see the spatially aggregated value.
From this, many different conclusion can be made, depending on what the user consider high or low densities.
In any case, the judgment will not reflect a density that actually exists anywhere.
The indicator on the other hand, can have different reference levels for each region, introducing additional ecological information about the carrying capacity of the ecosystems.
The levels may be such that the high density region is considered as having an overabundance of herbivores, and the low density region as having underabundance.
When normalised, the indicator value will be low for both regions, and the spatially aggregated value will also be low.
The indicator will describe the average condition for the intended ecosystem characteristic, whereas the variable will in essence not show anything that can be interpreted as such.

@@

As we have seen, aggregation bias causes a divergence in the type of information made available through the spatial aggregation of variables versus indicators.
In addition, unavoidable aggregation bias (for good and bad) will also affect the indicator values and the inferences that can be drawn from these.
If not careful, aggregation bias may also cause errors to sneak into the indicator logic.
A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition indicators (@fig-agg1; @fig-agg2), is the choice about the order of the different steps in the normalisation and aggregation process, i.e. the aggregation pathway (@fig-pathways).
If normalisation occurs early in the aggregation process, the amount of compensation is limited.
However, sometimes complete or partial compensation is justified, as in the example in @sec-pti, or as we saw in @sec-match where indicator development involved an initial step of spatially aggregating variables to align with the spatial configuration of reference levels (pathways 2 and 3; @fig-pathways).
Failure to harmonise the spatial scale between variables and reference levels before normalising the variable can lead to aggregation bias, as described in @sec-bill.

@@

Another cause of potential errors or confusion arising during the spatial aggregation of indicators has to do with the number of aggregation levels.
The SEEA EA uses a principle of commensurability, meaning that the order of certain operations should not affect the resulting estimate or value.
Spesifically, it should not matter whether you aggregated vertically first (to create an index) and then horizontally (i.e. spatial), or the other way around.
Note that this prisiple may not always hold for other cases, such as aggregating across nested areas, e.g. from ecosystem assets to EAAs, although this is perhaps considered desirable.
This has to do with the use of area weighting in the spatial aggregation.
Consider an indicator originating from point-based field surveys.
The sampling location are selected using a stratified protocol that ensures a representative sample of similar size (same *n*) is made for each ecosystem asset.
If the variable is normalised early in the aggregation pathway to create indicator values for each locality, then the mean of indicator values for each region will be representative for those regions, and no area weighting is required.
For the next spatial level, when information from all ecosystem assets should be used to create an estimate for the EAA, ecosystem assets with large extents should weight more in the calculations (following the SEEA EA prinsiples of area weighting).
Note that one cannot skip this first level of spatial aggregation via ecosystem assets, since this is an integral part of the sampling design, leading to spatial autocorrelation (se @sec-auto).
However, if in a later ECA, ecosystem assets are defined as subsets of the initial assets (a finer scale account), then the same indicator, when aggregated to the EAA level, will have a different value.
A similar case would be is sampling was random and dispersed across the EAA, lets say a country, and indicator values were first calculated as some average across all sampling points, and then in a later account or assessment, the same indicator was estimated for separate regions.
When aggregating the indicator values from regions to the whole country, using area weighting, the national estimate will differ, albeit slightly, from the initial national estimate that did not include this step-wise aggregation.
Cases when two accounts or assessments use what is considered the same indicator but find different values for the same areas, is cause for both confusion, and potentially mistrust.
In our experience this can also lead to time consuming detective work, trying to find the causes for such discrepancies.
We suggest two immediate solution to this issue.
First, to find better ways to document the aggregation pathways that are being used for each indicator, as part of the indicator metadata.
Second, to find ways to distinguish between apparently identical indicators, for example using different version numbers or IDs.
We provide practical solution to both these suggestions in @sec-rec.

>>>>

For sparse data we are considering data where variable values are assigned to singular points or small areas.
In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information.
To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some generalisations.
One can assign values to areas via an aggregation functions used on data values that fall within a given ecosystem asset.
Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54] .
When doing this aggregation, we need to account for the spatial structure (spatial correlations) of the data.
One way is to rely on design-based approaches (relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets.
In our experience, the choice of which area units to use when doing this step-wise aggregation is often not reflected much upon.
In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
However, the arbitrariness of this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; @nte-rel).
It is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
The MAUP can, however, lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].

@@

```{r fig-agg1}
#| echo: false
#| fig-cap: |
#|   Examples showing the difference in indicator values (y-axis) obtained from taking the mean of two variable values (V1 and V1) that are on the original scale (i hat) or on the normalised indicator scale (i bar). Pane A: Due to truncation at X~100~, the latter results in a comparatively lower indicator value. Pane B: Example where the normalisation function includes a transformation step (see Figure 1 for definition) that anchors X~60~ to 0.6 on the indicator scale. This reverses the order of i hat and i bar compared to example in pane A. The solid coloured line represents the normalisation function f(V). V = variable value, i = indicator value. Modified from Rastsetter (1991).
#| out-width: 600

knitr::include_graphics("../img/aggregation_bias.PNG")

@@

@@

In ecosystem condition accounting the aggregation of both variables and indicators is common practice and raises the issue of **aggregation bias** (@fig-agg1) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
This means that the spatial aggregation of metrics leads to a divergence in the type of information held in variables and in their associated indicators, to a point where they become qualitatively very different.
As an example, consider an indicator based on spatial information on the abundance of a large herbivore.
The ecosystem characteristic that this metric is meant to describe is "*Appropriate herbivore pressure, balancing overgrazing and shrubification"*.
The variable shows relatively high densities in half the area and low densities in the other half.
When spatially aggregating the variable, either by taking the mean density or the sum of individuals, will produce an estimate that is somewhere in the middle of the two regions.
The user will have no information about the carrying capacity for each of the regions, and will only see the spatially aggregated value.
From this, many different conclusion can be made, depending on what the user consider high or low densities.
In any case, the judgment will not reflect a density that actually exists anywhere.
The indicator on the other hand, can have different reference levels for each region, introducing additional ecological information about the carrying capacity of the ecosystems.
The levels may be such that the high density region is considered as having an overabundance of herbivores, and the low density region as having underabundance.
When normalised, the indicator value will be low for both regions, and the spatially aggregated value will also be low.
The indicator will describe the average condition for the intended ecosystem characteristic, whereas the variable will in essence not show anything that can be interpreted as such.

@@

As we have seen, aggregation bias causes a divergence in the type of information made available through the spatial aggregation of variables versus indicators.
In addition, unavoidable aggregation bias (for good and bad) will also affect the indicator values and the inferences that can be drawn from these.
If not careful, aggregation bias may also cause errors to sneak into the indicator logic.
A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition indicators (@fig-agg1), is the choice about the order of the different steps in the normalisation and aggregation process, i.e. the aggregation pathway (@fig-pathways).
If normalisation occurs early in the aggregation process, the amount of compensation is limited.
However, sometimes complete or partial compensation is justified, as in the example in @sec-pti, or as we saw in @sec-match where indicator development involved an initial step of spatially aggregating variables to align with the spatial configuration of reference levels (pathways 2 and 3; @fig-pathways).
Failure to harmonise the spatial scale between variables and reference levels before normalising the variable can lead to aggregation bias, as described in @sec-bill.

@@

Another cause of potential errors or confusion arising during the spatial aggregation of indicators has to do with the number of aggregation levels.
The SEEA EA uses a principle of commensurability, meaning that the order of certain operations should not affect the resulting estimate or value.
Spesifically, it should not matter whether you aggregated vertically first (to create an index) and then horizontally (i.e. spatial), or the other way around.
Note that this prisiple may not always hold for other cases, such as aggregating across nested areas, e.g. from ecosystem assets to EAAs, although this is perhaps considered desirable.
This has to do with the use of area weighting in the spatial aggregation.
Consider an indicator originating from point-based field surveys.
The sampling location are selected using a stratified protocol that ensures a representative sample of similar size (same *n*) is made for each ecosystem asset.
If the variable is normalised early in the aggregation pathway to create indicator values for each locality, then the mean of indicator values for each region will be representative for those regions, and no area weighting is required.
For the next spatial level, when information from all ecosystem assets should be used to create an estimate for the EAA, ecosystem assets with large extents should weight more in the calculations (following the SEEA EA prinsiples of area weighting).
Note that one cannot skip this first level of spatial aggregation via ecosystem assets, since this is an integral part of the sampling design, leading to spatial autocorrelation.
However, if in a later ECA, ecosystem assets are defined as subsets of the initial assets (a finer scale account), then the same indicator, when aggregated to the EAA level, will have a different value.
A similar case would be is sampling was random and dispersed across the EAA, lets say a country, and indicator values were first calculated as some average across all sampling points, and then in a later account or assessment, the same indicator was estimated for separate regions.
When aggregating the indicator values from regions to the whole country, using area weighting, the national estimate will differ, albeit slightly, from the initial national estimate that did not include this step-wise aggregation.
Cases when two accounts or assessments use what is considered the same indicator but find different values for the same areas, is cause for both confusion, and potentially mistrust.
In our experience this can also lead to time consuming detective work, trying to find the causes for such discrepancies.
We suggest two immediate solution to this issue.
First, to find better ways to document the aggregation pathways that are being used for each indicator, as part of the indicator metadata.
Second, to find ways to distinguish between apparently identical indicators, for example using different version numbers or IDs.
We provide practical solution to both these suggestions in @sec-rec.
